---
title: "Setting up a simple pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up a simple pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rixpress)
```

This vignette will introduce some jargon and walk you through setting up a
simple pipeline using `{rixpress}`.

## Definition

In Nix jargon, a *derivation* is *a specification for running an executable on
precisely defined input files to repeatably produce output files at uniquely
determined file system paths.*
([source](https://nix.dev/manual/nix/2.25/language/derivations))

In simpler terms, a derivation is a recipe with precisely defined inputs, steps,
and a fixed output, meaning, that for exactly the same inputs and exactly the 
same build steps, *exactly* the same output is produced. This is important to
understand, because to be able to always build exactly the same output, a number
of measures have to be take:

- a derivation must have all of its inputs declared explicitely;
- inputs are everything that plays a role in building an output:
  software dependencies, but also configuration flags or environment variables;
- to ensure the exact same output is always built,
  the build process happens in an *hermetic* sandbox.

This last point is quite important: because the build process must happen inside
a sandbox, if your building your output requires say, Quarto, Quarto must then
be explicitely listed as an input, *even* if you already have Quarto installed
on your system. The same goes for an environmental variable; for example, R
users may have to set the variable `JAVA_HOME` to make R aware of where the java
is installed. However, if you require java for a derivation
