---
title: "Setting Up a Pipeline From Scratch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting Up a Pipeline From Scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates how to build a `{rixpress}` pipeline that only uses R
derivations. For multilingual pipelines, read the next vignette
`vignette("polyglot")`.

If you are more of a visual learner, you can also watch
[this video](https://www.youtube.com/watch?v=IXKd5ySzzSU) which
will walk you through the contents of this vignette.

# Installing the required software

If you're using `{rixpress}`, you're likely already familiar with its sister
package `{rix}`, which helps you set up reproducible development environments
for R or Python using `Nix`. As such, you likely already have `Nix` installed on your
system. If not, read the following [`{rix}` vignette to learn
how to set up `Nix` on Linux or
Windows](https://docs.ropensci.org/rix/articles/installing-r-packages.html),
or [this one if you're on macOS](https://docs.ropensci.org/rix/articles/setting-up-macos.html).

Once `Nix` is installed (as well as `cachix`, and the `rstats-on-nix` cache
configured), you're ready to go.

# Bootstrapping a project

In this vignette, I'm going to assume the following:

- you're already familiar with `Nix` thanks to `{rix}`;
- you only use `Nix` to manage R environments for data science projects (in other
  words, you don't have a system-installed R);
- you have `Nix` installed on your system as well as `cachix` and configured the
  `rstats-on-nix` cache;
- you've read through the previous two vignettes.

If you're already familiar with `{rix}`, you likely know that you can use the
following command in your project's root directory to get access to a temporary
`Nix` shell that provides R, `{rix}`, and `{rixpress}`:

```bash
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
```

This will start a `Nix` session. Type `R` to start an R session, and then type:

```{r, eval = FALSE}
library(rixpress)

rxp_init()
```

This will write two files to the root of your project, `gen-env.R` and
`gen-pipeline.R`. `gen-env.R` will look like this:

```{r, eval = FALSE}
# This script defines the default environment the pipeline runs in.
# Add the required packages to execute the code necessary for each derivation.
# If you want to create visual representations of the pipeline, consider adding
# `{visNetwork}` and `{ggdag}` to the list of R packages.
library(rix)

# Define execution environment
rix(
  date = NULL,
  r_pkgs = NULL,
  py_conf = NULL,
  git_pkgs = list(
    "package_name" = "rixpress",
    "repo_url" = "https://github.com/ropensci/rixpress",
    "commit" = "HEAD"
  ),
  ide = "none",
  project_path = "."
)
```

To execute a `{rixpress}` pipeline, the environment needs `{rixpress}` to be
available, but you likely require other packages. Let's add `{dplyr}` and
`{igraph}` (which is needed to view the graphical representation of the
pipeline's DAG). Open `gen-env.R` in any text editor and change it to:

```{r, eval = FALSE}
library(rix)

# Define execution environment
rix(
  date = "2025-04-11",
  r_pkgs = c("dplyr", "igraph"),
  git_pkgs = list(
    package_name = "rixpress",
    repo_url = "https://github.com/ropensci/rixpress",
    commit = "HEAD"
  ),
  ide = "rstudio",
  project_path = ".",
  overwrite = TRUE
)
```

I've added a date, which means that the versions of R and R packages will be set
to the versions available on this date from CRAN, and I've added `{dplyr}` and
`{igraph}`. I've also added RStudio as the IDE. I can now go back to my
temporary `Nix` shell and run `Rscript gen-env.R` (or `source("gen-env.R")`
inside an R session). This will create a `default.nix` file in my project's
folder: this is a `Nix` file that defines my environment. I can now leave my
temporary shell by first quitting R (if applicable), then typing CTRL-D or
`exit`. Then I can build my development environment using `nix-build`. Let's now
enter our development shell by typing `nix-shell` and launch RStudio (or
whatever editor you want to use) and open `gen-pipeline.R`:


```{r, eval = FALSE}
library(rixpress)
library(igraph)

list(
  rxp_r_file(
    name = NULL,
    path = NULL,
    read_function = \(x) read.csv(file = x, sep = ",")
  ),
  rxp_r(
    name = NULL,
    expr = NULL
  )
) |> rxp_populate(build = TRUE)
```

This template starts with `rxp_r_file()` to load data, and `rxp_r()` to
define a step. These two derivations are put into a list and then passed to
`rxp_populate()`. Let's start by importing data and filtering it using `{dplyr}`:

```{r, eval = FALSE}
library(rixpress)
library(igraph)

list(
  rxp_r_file(
    name = mtcars,
    path = 'data/mtcars.csv',
    read_function = \(x) (read.csv(file = x, sep = "|"))
  ),

  rxp_r(
    name = filtered_mtcars,
    expr = filter(mtcars, am == 1)
  )
) |> rxp_populate(build = TRUE)
```

Select the pipeline and run it (I highly recommend you use keyboard shortcuts
to quickly run the above code when working interactively). If everything
went well, you will see the following in your R console:

```r
Build process started...

+ > mtcars building
+ > filtered_mtcars building
✓ filtered_mtcars built
✓ mtcars built
✓ pipeline completed [2 completed, 0 errored]
Build successful! Run `rxp_inspect()` for a summary.
Read individual derivations using `rxp_read()` or
load them into the global environment using `rxp_load()`.
```

Let’s inspect the pipeline outputs using `rxp_inspect()`:

```r
       derivation build_success
1 all-derivations          TRUE
2 filtered_mtcars          TRUE
3          mtcars          TRUE
                                                         path       output
1 /nix/store/xc67xyqw8ziggwdk7hhiqsgh7hxfv4k1-all-derivations filtered....
2 /nix/store/njv7zpgiq9p843dixkcg1dpryr5rn6h4-filtered_mtcars filtered....
3          /nix/store/gqf4l5jpbjdv1by0vw6vphhlrfa0600a-mtcars       mtcars
  error_message
1          <NA>
2          <NA>
3          <NA>
```

Three things were built: `mtcars`, which is simply the source data set;
`filtered_mtcars`, which is the filtered data; and an object you didn't
define called `all-derivations`. This last object is mostly for internal
`{rixpress}` use, and you can safely ignore it.

Let's now use `rxp_read()` to take a look at `filtered_mtcars`:

```{r, eval = FALSE}
rxp_read("filtered_mtcars")
```

```r
                mpg cyl  disp  hp drat    wt  qsec vs am gear carb
Mazda RX4      21.0   6 160.0 110 3.90 2.620 16.46  0  1    4    4
Mazda RX4 Wag  21.0   6 160.0 110 3.90 2.875 17.02  0  1    4    4
Datsun 710     22.8   4 108.0  93 3.85 2.320 18.61  1  1    4    1
Fiat 128       32.4   4  78.7  66 4.08 2.200 19.47  1  1    4    1
Honda Civic    30.4   4  75.7  52 4.93 1.615 18.52  1  1    4    2
Toyota Corolla 33.9   4  71.1  65 4.22 1.835 19.90  1  1    4    1
Fiat X1-9      27.3   4  79.0  66 4.08 1.935 18.90  1  1    4    1
Porsche 914-2  26.0   4 120.3  91 4.43 2.140 16.70  0  1    5    2
Lotus Europa   30.4   4  95.1 113 3.77 1.513 16.90  1  1    5    2
Ford Pantera L 15.8   8 351.0 264 4.22 3.170 14.50  0  1    5    4
Ferrari Dino   19.7   6 145.0 175 3.62 2.770 15.50  0  1    5    6
Maserati Bora  15.0   8 301.0 335 3.54 3.570 14.60  0  1    5    8
Volvo 142E     21.4   4 121.0 109 4.11 2.780 18.60  1  1    4    2
```

You can save this object like so:

```{r, eval = FALSE}
filtered_mtcars <- rxp_read("filtered_mtcars")
```

or you can simply use `rxp_load("filtered_mtcars")` which will achieve the same
result.

You can now manipulate this object interactively, as usual. You might still not
quite know what the next step should be (maybe there are missing values you need
to handle somehow, or another such cleaning step), but once you know the next
steps, you can write the next derivation:

```{r, eval = FALSE}
library(rixpress)
library(igraph)

list(
  rxp_r_file(
    name = mtcars,
    path = 'data/mtcars.csv',
    read_function = \(x) (read.csv(file = x, sep = "|"))
  ),

  rxp_r(
    name = filtered_mtcars,
    expr = filter(mtcars, am == 1)
  ),

  rxp_r(
    name = mtcars_mpg,
    expr = select(filtered_mtcars, mpg)
  )
) |>
  rxp_populate(build = TRUE)
```

Rebuild the pipeline. The first two steps are skipped because they're already
available in the `Nix` store. You can now start the *inspect -> try stuff ->
define derivation -> build pipeline* loop again for the next derivation.

You now know the basics of using `{rixpress}` to build a simple pipeline.

The next vignette `vignette("polyglot")` explains how to build a pipeline
with both R and Python derivations, as well as compiling a Quarto document.

If you don't need to work with multiple languages, you can continue with
the vignette `vignette("debugging")`.

